---
title: "Comparing Two pathfindR Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Two pathfindR Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, fig.height = 4, fig.align = "center"
)
suppressPackageStartupMessages(library(pathfindR))
```

The function `combine_pathfindR_results()` allows combination of two pathfindR active-subnetwork-oriented enrichment analysis results for investigating common and distinct terms between the groups. Below is an example for comparing results using two different rheumatoid arthritis-related data sets (`RA_output` and `RA_comparison_output`).

```{r compare2res}
combined_df <- combine_pathfindR_results(result_A = RA_output, 
                                         result_B = RA_comparison_output, 
                                         plot_common = FALSE)
```

By default, `combine_pathfindR_results()` plots the term-gene graph for the common terms in the combined results. For not plotting the graph, set `plot_common = FALSE`. 

The function `combined_results_graph()` can be used to create this graph (using only selected terms etc.) later on. By default, the function creates the graph using all common terms:

```{r compare_graph1}
combined_results_graph(combined_df)
```

By supplying a vector of selected terms to the `selected_terms` arguments, you may plot the term-gene graph for the selected terms:

```{r compare_graph2, fig.width=8, fig.height=4}
combined_results_graph(combined_df, 
                       selected_terms = c("hsa04144", "hsa04141", "hsa04140"))
```

By default, `combined_results_graph()` creates the graph using term IDs. To use term descriptions instead, set `use_description = TRUE`: 

```{r compare_graph3, eval=FALSE}
combined_results_graph(combined_df, 
                       use_description = TRUE,
                       selected_terms = combined_df$Term_Description[1:4])
```

For changing the layout of the graph (`"auto"` by default), you may use the `layout` argument.

For changing how the sizes of the term nodes are determined, you may use the `node_size` argument. The options are `"num_genes"` (default) and `"p_val"` for using the number of significant genes in the term and the -log10(p) value of the term, respectively:

```{r compare_graph4, eval=FALSE}
combined_results_graph(combined_df, 
                       selected_terms = c("hsa04144", "hsa04141", "hsa04140"),
                       node_size = "p_val")
```
